---
title: "Analysis"
date: "`r Sys.Date()`"
output: pdf_document
header-includes:
  - \usepackage{array}

---

```{r setup, include=FALSE}
library(kyotil)
library(knitr)
library(kableExtra)
library(lme4)
library(readxl)
library(glue)
library(dplyr)

# global variables, don't redefine in each section
cytokines=c("CXCL10", "CCL5",  "IL15",  "IL12A", "IFI16"); names(cytokines)=cytokines
trts=c("Mock", "100U", "Virus", "Virus+100U"); names(trts)=trts
donors=c(4,5,7,10,1,2,3,6,8,9)
doses=c("0", "0.1", "1", "10", "100"); names(doses)=doses
tissues=c("Keratinocytes", "Fibroblasts"); names(tissues)=tissues
times=c("Untreated", "4HPT", "24HPT"); names(times)=times
times2=c("30M", "60M", "120M", "240M"); names(times2)=times2


f.paired.t = function(dat, outcome, cond, print=F) {

  # take average of replicates for each donor and trt
  tmp = dat %>%
    group_by(donor, !!sym(cond)) %>%
    summarize(fc = mean(.data[[outcome]], na.rm = TRUE), .groups = "drop")

  # reshape to wide format with one column per trt
  tmp1 = reshape(as.data.frame(tmp), idvar = "donor", timevar = cond,
                 direction = "wide")

  # extract the two treatment groups inside the cond column
  groups <- unique(dat[[cond]])
  if (length(groups) != 2) {
    stop("The cond column must contain exactly 2 treatment levels.")
  }
  # rename for clarity
  g1 <- groups[1]
  g2 <- groups[2]

  # print(tmp1)
  
  # run paired t-test
  test = t.test(tmp1[["fc."%.%g1]], tmp1[["fc."%.%g2]], paired = TRUE)

  if (print) {
    myboxplot(list(tmp1[["fc."%.%g1]], tmp1[["fc."%.%g2]]))
    print(tmp1)
    print(test)
  }
  
  test$p.value
}

```

## Background 

This project focuses on cytokines and their role in controlling viral replication within the epithelium. 

## Methods

Fold changes are computed from actin.

When data are paired, paired t tests are used. When data are not paired, both unpaired t tests and linear mixed models (lmm) are used to analyze the data. t tests first average the outcome in the replicates, lmm uses all data. Lmm may be more efficient, but when there are warnings about singular values, the validyt of their results is in doubt. 

\newpage
## Fig 5A

Data is log10 transformed.


```{r panel A data, echo=F, message=T}

dat=read.csv("data/Panel5A.csv")

# reshape to long format
dat.1 = list(dat[, c(1,2,3:5)], dat[, c(1,2,3:5+3)], dat[, c(1,2,3:5+6)], dat[, c(1,2,3:5+9)],
             dat[, c(1,2,3:5+12)], dat[, c(1,2,3:5+15)], dat[, c(1,2,3:5+18)], dat[, c(1,2,3:5+21)],
             dat[, c(1,2,3:5+24)], dat[, c(1,2,3:5+27)])
# add a donor column to each data frame
dat.1 = lapply (seq_along(donors), function (i) {
  x=dat.1[[i]]
  x$donor=donors[i]
  # reshape to long format for columns 3-5
  x=myreshapelong(x, cols.to.be.stacked=names(x)[3:5], label.cols.to.be.stacked="replicate", new.col.name="fold_change")
  rownames(x)=NULL
  x$id=NULL
  # change replicate to 2 if ends with .1, 3 if ends with .2, 1 otherwise
  x$replicate=ifelse(grepl("\\.1$", x$replicate), 2, ifelse(grepl("\\.2$", x$replicate), 3, 1))
  x
})

do.call(rbind, dat.1) -> dat.2
# remove reshapeLong attr
attr(dat.2, "reshapeLong")=NULL

# add a new column tissue = "keratinocyte" if exp starts with K or "fibroblast" if exp starts with F
dat.2$tissue=ifelse(grepl("^K", dat.2$exp), "keratinocyte", "fibroblast")
# add a new column trt = "treated" if exp ends with T or "untreated" if exp ends with U
dat.2$trt=ifelse(grepl("T$", dat.2$exp), "treated", "untreated")
# remove exp column
dat.2$exp=NULL

# make tissue, trt, cytokine, donor factors
dat.2$tissue=factor(dat.2$tissue)
dat.2$trt=factor(dat.2$trt, levels=c("untreated", "treated"))
dat.2$cytokine=factor(dat.2$cytokine, levels=cytokines)
dat.2$donor=factor(dat.2$donor)
dat.2$replicate=factor(dat.2$replicate)

dat.2$fold_change=log10(dat.2$fold_change)

dat=dat.2
```

**Question**: Is there a statistically significant difference between the treated group and its associated untreated control? 

For Keratinocyte, there are 10 donors; for Fibroblast, there are 4 donors. 

```{r 5a1, echo=F, message=T}
mysapply(cytokines, function (c) {
  # # lmm
  # fit.1 = lmer(fold_change ~ trt + (1|donor), data=subset(dat, cytokine==c & tissue=="keratinocyte" & !is.na(fold_change)))
  # fit.2 = lmer(fold_change ~ trt + (1|donor), data=subset(dat, cytokine==c & tissue=="fibroblast" & !is.na(fold_change)))
  # tab=getFormattedSummary(list(Keratinocyte=fit.1, Fibroblast=fit.2), robust=F, p.digits=3, exp10=T, trunc.large.est = F)
  # tab[2,]
  
  # paired t test
  out = c(
    Keratinocyte=f.paired.t(subset(dat, cytokine==c & tissue=="keratinocyte" & !is.na(fold_change)), outcome="fold_change", cond="trt"),
    Fibroblast  =f.paired.t(subset(dat, cytokine==c & tissue=="fibroblast"   & !is.na(fold_change)), outcome="fold_change", cond="trt"))
  formatPvalues(out)
}) -> tab
kable(tab, align='l',
      caption="Paired t test p values.") %>%
  kable_styling(latex_options = "hold_position")%>% 
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[4pt]")
```
Treated versus untreated is highly significant except for CCL5 on fibroblasts.

\newpage

**Question**: Is there a statistically significant difference between treated keratinocytes and treated fibroblasts

```{r 5a2, echo=F, message=T}
mysapply(cytokines, function (c) {
  # # lmm
  # fit.1 = lmer(fold_change ~ tissue + (1|donor), data=subset(dat, cytokine==c & trt=="treated"))
  # fit.2 = lmer(fold_change ~ tissue + (1|donor), data=subset(dat, cytokine==c & trt=="untreated"))
  # tab=getFormattedSummary(list(Treated=fit.1, Untreated=fit.2), robust=F, p.digits=3, exp10=T, trunc.large.est = F)
  # tab[2,]
  
  # paired t test
  out = c(
    treated  =f.paired.t(subset(dat, cytokine==c & trt=="treated"   & !is.na(fold_change)), "fold_change", "tissue"),
    untreated=f.paired.t(subset(dat, cytokine==c & trt=="untreated" & !is.na(fold_change)), "fold_change", "tissue")
  )
  formatPvalues(out)
  
}) -> tab
kable(tab, align='l', 
      caption = "Paired t test p values.") %>%
  kable_styling(latex_options = "hold_position")%>% 
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[4pt]")
```
Treated keratinocytes versus treated fibroblasts is highly significant except for CCL5. Untreated keratinocytes versus untreated fibroblasts is not significant for any cytokine.


\newpage

## Fig 5B

Data is not log10 transformed.

```{r 5B data, echo=F, message=T}

dat=read.csv("data/Panel5B.csv")

# reshape to long format
dat.1 = list(dat[, c(1:3,4:6)], dat[, c(1:3,4:6+3)], dat[, c(1:3,4:6+6)], dat[, c(1:3,4:6+9)],
             dat[, c(1:3,4:6+12)], dat[, c(1:3,4:6+15)], dat[, c(1:3,4:6+18)], dat[, c(1:3,4:6+21)],
             dat[, c(1:3,4:6+24)], dat[, c(1:3,4:6+27)])
# add a donor column to each data frame
dat.1 = lapply (seq_along(donors), function (i) {
  x=dat.1[[i]]
  x$donor=donors[i]
  # reshape to long format for columns 4-6
  x=myreshapelong(x, cols.to.be.stacked=names(x)[4:6], label.cols.to.be.stacked="replicate", new.col.name="dCt")
  rownames(x)=NULL
  x$id=NULL
  # change replicate to 2 if ends with .1, 3 if ends with .2, 1 otherwise
  x$replicate=ifelse(grepl("\\.1$", x$replicate), 2, ifelse(grepl("\\.2$", x$replicate), 3, 1))
  x
})

do.call(rbind, dat.1) -> dat.2
# remove reshapeLong attr
attr(dat.2, "reshapeLong")=NULL

# make tissue, trt, cytokine, donor factors
dat.2$tissue=factor(dat.2$tissue)
dat.2$dose=factor(dat.2$dose, levels=doses)
dat.2$cytokine=factor(dat.2$cytokine, levels=cytokines)
dat.2$donor=factor(dat.2$donor)
dat.2$replicate=factor(dat.2$replicate)

dat=dat.2

```

**Question**: Is there a statistically significant difference between keratinocytes (orange) and fibroblasts (blue) at each dose in each cytokine of interest?

```{r 5b1, echo=F, message=T}
mysapply(cytokines, function (c) {
  # # lmm
  # fits = lapply(doses, function(d) lmer(dCt ~ tissue + (1|donor), data=subset(dat, cytokine==c & dose==d)))
  # names(fits)=doses
  # tab=getFormattedSummary(fits, robust=F, p.digits=3, exp10=F, trunc.large.est = F)
  # tab[2,]
  
  # paired t test
  out=sapply(doses, function(d) f.paired.t(subset(dat, cytokine==c & dose==d & !is.na(dCt)), "dCt", "tissue"))
  formatPvalues(out)

}) -> tab
kable(tab, align='c',
      caption = "Paired t test p values.") %>%
  kable_styling(latex_options = "hold_position") %>% 
  row_spec(0, bold = TRUE, align = "c")%>%
  column_spec(1, width = "1.1cm") %>%  
  column_spec(2:(ncol(tab)+1), width = "3cm")%>%  
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")
```

\newpage

**Question**: For each tissue, is there a statistically significant difference between each dose and the next lower dose? IE is there a difference between 0.1 and 1, 1 and 10, 10 and 100? The goal of this question is to establish what dose is the peak response

```{r 5b2, echo=F, message=T}
tabs=list()
for (j in 1:2) {
  tabs[[j]] = mysapply(cytokines, function (c) {
    # # lmm
    # fits = lapply(seq_along(doses)[-1], function(i) lmer(dCt ~ dose + (1|donor), data=subset(dat, tissue==tissues[j] & cytokine==c & dose %in% doses[c(i-1,i)])))
    # names(fits)=paste0(doses[2:length(doses)], " vs. ", doses[2:length(doses)-1])
    # tab=getFormattedSummary(fits, robust=F, p.digits=3, exp10=F, trunc.large.est = F)
    # tab[2,]
    
    # paired t test
    out = sapply(seq_along(doses)[-1], function(i) {
      tmp=subset(dat, tissue==tissues[j] & cytokine==c & dose %in% doses[c(i-1,i)])
      f.paired.t(tmp, "dCt", "dose")
    })
    names(out)=paste0(doses[2:length(doses)], " vs. ", doses[2:length(doses)-1])
    formatPvalues(out)
  }) 
}
# combine the two tables by interspering the rows
tab=do.call(rbind, lapply(1:nrow(tabs[[1]]), function(i) {
  rbind(tabs[[1]][i,], tabs[[2]][i,])
}))
rownames(tab)=c(t(outer(cytokines, tissues, paste, sep=", ")))
kable(tab, align="c",
      caption = "Paired t test p values.") %>%
  kable_styling(latex_options = "hold_position") %>% 
  row_spec(0, bold = TRUE, align = "c")%>%
  column_spec(1, width = "2.5cm") %>%  
  column_spec(2:(ncol(tab)+1), width = "3cm")%>%  
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")
```



\newpage
## Fig 5C

Data is log10 transformed.


```{r 5C data, echo=F, message=T}

dat=read.csv("data/Panel5C.csv")

# reshape to long format
dat.1 = list(dat[, c(1:3,4:6)], dat[, c(1:3,4:6+3)], dat[, c(1:3,4:6+6)], dat[, c(1:3,4:6+9)],
             dat[, c(1:3,4:6+12)], dat[, c(1:3,4:6+15)], dat[, c(1:3,4:6+18)], dat[, c(1:3,4:6+21)],
             dat[, c(1:3,4:6+24)], dat[, c(1:3,4:6+27)])
# add a donor column to each data frame
dat.1 = lapply (seq_along(donors), function (i) {
  x=dat.1[[i]]
  x$donor=donors[i]
  # reshape to long format for columns 4-6
  x=myreshapelong(x, cols.to.be.stacked=names(x)[4:6], label.cols.to.be.stacked="replicate", new.col.name="fold_change")
  rownames(x)=NULL
  x$id=NULL
  # change replicate to 2 if ends with .1, 3 if ends with .2, 1 otherwise
  x$replicate=ifelse(grepl("\\.1$", x$replicate), 2, ifelse(grepl("\\.2$", x$replicate), 3, 1))
  x
})

do.call(rbind, dat.1) -> dat.2
# remove reshapeLong attr
attr(dat.2, "reshapeLong")=NULL

# make tissue, trt, cytokine, donor factors
dat.2$tissue=factor(dat.2$tissue)
dat.2$dose=factor(dat.2$dose, levels=doses)
dat.2$cytokine=factor(dat.2$cytokine, levels=cytokines)
dat.2$donor=factor(dat.2$donor)
dat.2$replicate=factor(dat.2$replicate)

dat.2$fold_change=log10(dat.2$fold_change)

dat=dat.2
```

**Question**: Is there a statistically significant difference between keratinocytes (orange) and fibroblasts (blue) at each dose in each cytokine of interest?

```{r 5c1, echo=F, message=T}
mysapply(cytokines, function (c) {
  # # lmm
  # fits = lapply(doses, function(d) lmer(fold_change ~ tissue + (1|donor), data=subset(dat, cytokine==c & dose==d)))
  # names(fits)=doses
  # tab=getFormattedSummary(fits, robust=F, p.digits=3, exp10=T, trunc.large.est = F)
  # tab[2,]
  
  # paired t test
  out=sapply(doses, function(d) f.paired.t(subset(dat, cytokine==c & dose==d & !is.na(fold_change)), "fold_change", "tissue"))
  formatPvalues(out)
  
}) -> tab

kable(tab, align='c',
      caption = "Paired t test p values.") %>%
  kable_styling(latex_options = "hold_position") %>% 
  row_spec(0, bold = TRUE, align = "c")%>%
  column_spec(1, width = "1.1cm") %>%  
  column_spec(2:(ncol(tab)+1), width = "3cm")%>%  
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")
```

\newpage

**Question**: For each tissue, is there a statistically significant difference between each dose and the next lower dose? The goal of this question is to establish what dose is the peak response

```{r 5c2, echo=F, message=T}
tabs=list()
for (j in 1:2) {
  tabs[[j]] = mysapply(cytokines, function (c) {
    # # lmm
    # fits = lapply(seq_along(doses)[-1], function(i) lmer(fold_change ~ dose + (1|donor), data=subset(dat, tissue==tissues[j] & cytokine==c & dose %in% doses[c(i-1,i)])))
    # names(fits)=paste0(doses[2:length(doses)], " vs. ", doses[2:length(doses)-1])
    # tab=getFormattedSummary(fits, robust=F, p.digits=3, exp10=T, trunc.large.est = F)
    # tab[2,]
    
    # paired t test
    out = sapply(seq_along(doses)[-1], function(i) {
      tmp=subset(dat, tissue==tissues[j] & cytokine==c & dose %in% doses[c(i-1,i)])
      f.paired.t(tmp, "fold_change", "dose")
    })
    names(out)=paste0(doses[2:length(doses)], " vs. ", doses[2:length(doses)-1])
    formatPvalues(out)
  }) 
}
# combine the two tables by interspering the rows
tab=do.call(rbind, lapply(1:nrow(tabs[[1]]), function(i) {
  rbind(tabs[[1]][i,], tabs[[2]][i,])
}))
rownames(tab)=c(t(outer(cytokines, tissues, paste, sep=", ")))
kable(tab, align='c',
      caption = "Paired t test p values.") %>%
  kable_styling(latex_options = "hold_position") %>% 
  row_spec(0, bold = TRUE, align = "c")%>%
  column_spec(1, width = "2.5cm") %>%  
  column_spec(2:(ncol(tab)+1), width = "3cm")%>%  
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[5pt]")

```



\newpage
## Fig 5D

Data is log10 transformed.


```{r 5D data, echo=F, message=T}

dat=read.csv("data/Panel5D.csv")

# reshape to long format
dat.1 = list(dat[, c(1:3,4:6)], dat[, c(1:3,4:6+3)], dat[, c(1:3,4:6+6)], dat[, c(1:3,4:6+9)],
             dat[, c(1:3,4:6+12)], dat[, c(1:3,4:6+15)], dat[, c(1:3,4:6+18)], dat[, c(1:3,4:6+21)],
             dat[, c(1:3,4:6+24)], dat[, c(1:3,4:6+27)])
# add a donor column to each data frame
dat.1 = lapply (seq_along(donors), function (i) {
  x=dat.1[[i]]
  x$donor=donors[i]
  # reshape to long format for columns 4-6
  x=myreshapelong(x, cols.to.be.stacked=names(x)[4:6], label.cols.to.be.stacked="replicate", new.col.name="fold_change")
  rownames(x)=NULL
  x$id=NULL
  # change replicate to 2 if ends with .1, 3 if ends with .2, 1 otherwise
  x$replicate=ifelse(grepl("\\.1$", x$replicate), 2, ifelse(grepl("\\.2$", x$replicate), 3, 1))
  x
})

do.call(rbind, dat.1) -> dat.2
# remove reshapeLong attr
attr(dat.2, "reshapeLong")=NULL

# make tissue, trt, cytokine, donor factors
dat.2$tissue=factor(dat.2$tissue)
dat.2$time=factor(dat.2$time, levels=times)
dat.2$cytokine=factor(dat.2$cytokine, levels=cytokines)
dat.2$donor=factor(dat.2$donor)
dat.2$replicate=factor(dat.2$replicate)

dat.2$fold_change=log10(dat.2$fold_change)

dat=dat.2
```

**Question**: Is there a statistically significant difference between keratinocytes (orange) and fibroblasts (blue) at each time point for each cytokine individually? 

```{r 5D1, echo=F, message=T}
mysapply(cytokines, function (c) {
  # # lmm
  # fits = lapply(times, function(t) lmer(fold_change ~ tissue + (1|donor), data=subset(dat, cytokine==c & time==t)))
  # names(fits)=times
  # tab=getFormattedSummary(fits, robust=F, p.digits=3, exp10=T, trunc.large.est = F)
  # tab[2,]
  
  # paired t test
  out=sapply(times, function(t) f.paired.t(subset(dat, cytokine==c & time==t & !is.na(fold_change)), "fold_change", "tissue"))
  formatPvalues(out)
}) -> tab

kable(tab, align='c',
      caption = "Paired t test p values.") %>%
  kable_styling(latex_options = "hold_position") %>% 
  row_spec(0, bold = TRUE, align = "c")%>%
  column_spec(1, width = "1.1cm") %>%  
  column_spec(2:(ncol(tab)+1), width = "3cm")%>%  
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[5pt]")
```


**Question**: For fibroblasts, is there a statistically significant difference between untreated and 4HPT for each cytokine? Is there a statistically significant difference between 4HPT and 24HPT for each cytokine?

```{r 5D2, echo=F, message=T}
tabs=list()
tabs2=list()
for (j in 1:2) {
  tabs[[j]] = mysapply(cytokines, function (c) {
    # # lmm fit
    # fits = lapply(seq_along(times)[-1], function(i) {
    #   tmp=subset(dat, tissue==tissues[j] & cytokine==c & time %in% times[c(i-1,i)])
    #   lmer(fold_change ~ time + (1|donor), data=tmp)
    # })
    # names(fits)=paste0(times[2:length(times)], " vs. ", times[2:length(times)-1])
    # tab=getFormattedSummary(fits, robust=F, p.digits=3, exp10=T, trunc.large.est = F)
    # tab[2,]
    
    # paired t test
    pvals = sapply(seq_along(times)[-1], function(i) {
      tmp=subset(dat, tissue==tissues[j] & cytokine==c & time %in% times[c(i-1,i)] & !is.na(fold_change))
      f.paired.t(tmp, "fold_change", "time")
    })
    names(pvals)=paste0(times[2:length(times)], " vs. ", times[2:length(times)-1])
    formatPvalues(pvals)
  }) 
}

# combine the two tables by interspering the rows
tab=do.call(rbind, lapply(1:nrow(tabs[[1]]), function(i) {
  rbind(tabs[[1]][i,], tabs[[2]][i,])
}))
rownames(tab)=c(t(outer(cytokines, tissues, paste, sep=", ")))
kable(tab, align = 'c',
      caption = "Paired t test p values.") %>%
  kable_styling(latex_options = "hold_position") %>% 
  row_spec(0, bold = TRUE, align = "c")%>%
  column_spec(1, width = "2.5cm") %>%  
  column_spec(2:(ncol(tab)+1), width = "3cm")%>%  row_spec(1:9, extra_latex_after = "\\addlinespace[5pt]")
```



\newpage
## Fig 4C

Data is not log10 transformed.

```{r, message=F, echo=F}
dat1=readxl::read_xlsx("data/Data_CytokinePaper_IRH-AZ_2025-11-6.xlsx", sheet="Figure 4C")
```

```{r Fig 4C data, echo=F, message=T}
IC50=as.numeric(unlist(dat1[3,-1]))
dat=data.frame(donor=as.character(rep(c(1:4,  6,7,9,10), each=3)), IC50, severe=rep(c(0,1), each=12))

dat$donor=factor(dat$donor)

```

**Question**: Compare the IC50 value between asymptomatic and severe donors. Data represents three independent biological replicates for each donor

```{r 4C, echo=F, message=T}
# myboxplot(IC50~severe, dat, col.points=as.numeric(dat$donor), xlab="Severe", cex=1.2, pch=16)

fit = lmer(IC50 ~ severe + (1|donor), data=dat)
tab=getFormattedSummary(list(IC50=fit), robust=F, p.digits=3, trunc.large.est = F)
kable(tab[2,,drop=F], align='l', caption="Linear mixed model.")
```


\newpage
## Fig 4D

Data is not log10 transformed.

```{r, message=F, echo=F}
dat1=readxl::read_xlsx("data/Data_CytokinePaper_IRH-AZ_2025-11-6.xlsx", sheet="Figure 4D")
```

```{r Fig 4D data, echo=F, message=T}

IFNGR1=as.numeric(unlist(dat1[3,-1]))
STAT1=as.numeric(unlist(dat1[8,-1]))

dat.IFNGR1=data.frame(donor=as.character(rep(c(1:4,6,7,9,10), each=3)), IFNGR1, severe=rep(c(0,1), each=12))
dat.STAT1=data.frame(donor=as.character(rep(c(1:4,6,7,9,10), each=3)), STAT1, severe=rep(c(0,1), each=12))

dat.IFNGR1$donor=factor(dat.IFNGR1$donor)
dat.STAT1$donor=factor(dat.STAT1$donor)

```

**Question**: Compare the expression levels of both genes between asymptomatic and severe donors. Data represents three independent biological replicates for each donor


```{r IFNGR1, echo=F, message=T}
# myboxplot(IFNGR1~severe, dat.IFNGR1, col.points=as.numeric(dat.IFNGR1$donor), xlab="Severe", cex=1.2, pch=16)

fit = lmer(IFNGR1 ~ severe + (1|donor), data=dat.IFNGR1)
tab=getFormattedSummary(list(IFNGR1=fit), robust=F, p.digits=3, trunc.large.est = F)
kable(tab[2,,drop=F], align='l', caption="Linear mixed model.")
```


```{r STAT1, echo=F, message=T}
# myboxplot(STAT1~severe, dat.STAT1, col.points=as.numeric(dat.STAT1$donor), xlab="Severe", cex=1.2, pch=16)

fit = lmer(STAT1 ~ severe + (1|donor), data=dat.STAT1)
tab=getFormattedSummary(list(STAT1=fit), robust=F, p.digits=3, trunc.large.est = F)
kable(tab[2,,drop=F], align='l', caption="Linear mixed model.")
```



\newpage
## Fig 6A, 6B, and 6C

Data is log10 transformed for 6B and 6C and not 6A.

**Question**: Determine if there is a statistically significant difference between asymptomatic and severe donors at each dose (6A, 6B) or at each time (6C) for each gene (CXCL10, CCL5, IL15, IL12, IFI16) at dose 100. 

Data represents three independent biological replicates for all 10 donors. 

We first analyzed the data with linear mixed models. There are some warnings about singular values in 6B and 6C. Hence we re-analyzed the data with unpaired t tests.



```{r Fig 6ABC, echo=F, message=T, results='asis'}

genes=c("CXCL10", "CCL5",  "IL15",  "IL12A", "IFI16"); names(genes)=genes

for (a in c("A", "B", "C")) {
  
  dat1=suppressMessages(
    readxl::read_xlsx("data/Data_CytokinePaper_IRH-AZ_2025-11-6.xlsx", sheet=glue("Figure 6{a}"))
  )
  
  doOneRow=function (x) {
    res = data.frame(donor=as.character(rep(c(1:10), each=3)), 
               readout=(as.numeric(unlist(x[-(1:2)]))), 
               severe=rep(c(0,1), each=15))
    if (a=="B" | a=="C") {
      res$readout = log10(res$readout)
    }
    res
  }
  
  if (a=="A" | a=="B") {
    rows=c(CXCL10=list(4:8), CCL5=list(12:16), IL15=list(20:24), IL12A=list(28:32), IFI16=list(36:40))
  } else if (a=="C") {
    rows=c(CXCL10=list(4:6), CCL5=list(10:12), IL15=list(16:18), IL12A=list(22:24), IFI16=list(28:30))
  }
  
  tmp = if (a=="A" | a=="B") doses else times
  
  lapply(genes, function (gene) {
    res=lapply(rows[[gene]], function (i) doOneRow(dat1[i-1,]))
    names(res)= tmp
    res
  }) -> dat.list
  
  for (iter in 1:2) {
    mysapply(genes, function (gene) {
      # print(gene)
      if (iter==1) {
        # lmm
        fits <- lapply(tmp, function(d) {
        
          fit <- lmer(readout ~ severe + (1|donor), data = dat.list[[gene]][[d]])
        
          # flag warning if singular or has convergence messages
          warn_flag <- FALSE
        
          # singularity
          if (isSingular(fit, tol = 1e-4)) {
            warn_flag <- TRUE
          }
        
          # convergence warnings stored in optinfo
          msgs <- fit@optinfo$conv$lme4$messages
          if (!is.null(msgs)) {
            warn_flag <- TRUE
          }
        
          # store attribute
          attr(fit, "warning_occurred") <- warn_flag
          attr(fit, "convergence_messages") <- msgs
        
          fit
        })

        tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
        out = tab[2,]
        # check for warnings
        warnings = sapply(fits, function(fit) attr(fit, "warning_occurred"))
        # print(warnings)
        out[warnings] = paste0(out[warnings], "W")
        out
        
      } else {
        # unpaired t test
        pvals = sapply(tmp, function(d) {
          tmp1=dat.list[[gene]][[d]] %>%
            group_by(donor, severe) %>%
            summarize(readout=mean(readout), .groups="drop")
          # par(mfrow=c(1,2))
          # myboxplot(readout~severe, data=tmp1, ylim=c(-1,1))
          # myboxplot(readout~severe, data=tmp1)
          t.test(readout~severe, data=tmp1)$p.val
        })
        formatPvalues(pvals)
      }  
    }) -> tab
    tab
  
    tab=cbind(rownames(tab), tab)
    colnames(tab)[1]=glue("6{a}")
    print(
      kable(tab, align="c", row.names=FALSE, caption=ifelse(iter==1,"Linear mixed models.", "Unpaired t tests.")) %>%  
      kable_styling(latex_options = "hold_position") %>% 
      column_spec(1, width = "1.1cm") %>%  
      column_spec(2:(ncol(tab)), width = "3cm")%>%  # no 1+ncol b/c row.names is FALSE
      row_spec(0, bold = TRUE, align = "c")%>%
      row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")
    ) 
  }
  
  cat("\\newpage\n")   # page break in LaTeX

}
```

## Fig 7A and 7B


**Question**: Determine if there is a statistically significant difference between the expression level of each gene (CXCL10, CCL5, IL15, IL12, IFI16) at each dose compared to the previous dose. For clarity, this is comparing 0.0 to 0.1, 0.1 to 1, 1 to 10, and 10 to 100. Data represents three independent biological replicates from 10 different donors

Data is log10 transformed.

```{r Fig 7A, echo=F, message=T, results='asis'}

dat1=suppressMessages(
  readxl::read_xlsx("data/Data_CytokinePaper_IRH-AZ_2025-11-6.xlsx", sheet=glue("Figure 7A"))
)

doOneRow=function (x) {
  # use x[3:32] instead of x[-(1:2)] b/c there are some extra columns, not visible in excel, at the end
  data.frame(donor=as.character(rep(c(1:10), each=3)), readout=log10(as.numeric(unlist(x[3:32]))), severe=rep(c(0,1), each=15))
}

genes1=c("CXCL9", "CXCL10",  "CXCL11"); names(genes1)=genes1
rows=c("CXCL9"=list(4:8), "CXCL10"=list(12:16), "CXCL11"=list(20:24))

lapply(genes1, function (gene) {
  res=lapply(rows[[gene]], function (i) doOneRow(dat1[i-1,]))
  names(res)=doses
  res
}) -> dat.list

mysapply(genes1, function (gene) {
  # # linear mixed model
  # fits = lapply(2:len(doses), function(i) {
  #   dat.tmp = rbind(
  #     cbind(dose=doses[i-1], dat.list[[gene]][[doses[i-1]]], row.names=NULL),
  #     cbind(dose=doses[i],   dat.list[[gene]][[doses[i]]], row.names=NULL)
  #   )
  #   dat.tmp$dose=factor(dat.tmp$dose, levels=doses)
  # 
  #   # suppressWarnings(
  #     lmer(readout ~ dose + (1|donor), data=dat.tmp)
  #   # )
  # })
  # names(fits)=doses[-1]
  # tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
  # res = tab[2,]
  
  # paired t test
  out = sapply(2:len(doses), function(i) {
    dat.tmp = rbind(
      cbind(dose=doses[i-1], dat.list[[gene]][[doses[i-1]]], row.names=NULL),
      cbind(dose=doses[i],   dat.list[[gene]][[doses[i]]], row.names=NULL)
    )
    dat.tmp$dose=factor(dat.tmp$dose, levels=doses)
    f.paired.t(dat.tmp, "readout", "dose")
  })
  names(out)=paste0(doses[2:length(doses)], " vs. ", doses[2:length(doses)-1])
  res = formatPvalues(out)
    
  all_same = sapply(2:len(doses), function(i) {
    dat.tmp = rbind(
      cbind(dose=doses[i-1], dat.list[[gene]][[doses[i-1]]], row.names=NULL),
      cbind(dose=doses[i],   dat.list[[gene]][[doses[i]]], row.names=NULL)
    )
    len(unique(dat.tmp$readout))==1
  })
  res[all_same] = "all values identical"
  res
}) -> tab
# colnames(tab)=glue("{colnames(tab)} vs. Prev. Dose")

tab=cbind(rownames(tab), tab)
colnames(tab)[1]="7A"
kable(tab, align="c", caption="Paired t tests.", row.names=F)%>%  
  kable_styling(latex_options = "hold_position") %>% 
  # column_spec(1, width = "1.1cm") %>%  
  column_spec(1:(ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")


```

\newpage

**Question**: Determine if there is a statistically significant difference between asymptomatic and severe donors at each dose (0, 0.1, 1, 10, 100). Data represents three independent biological replicates for all 10 donors.

Data is log10 transformed.

```{r Fig 7B, echo=F, message=T, results='asis'}

dat1=suppressMessages(
  readxl::read_xlsx("data/Data_CytokinePaper_IRH-AZ_2025-11-6.xlsx", sheet=glue("Figure 7B"))
)

doOneRow=function (x) {
  # use x[3:32] instead of x[-(1:2)] b/c there are some extra columns, not visible in excel, at the end
  data.frame(donor=as.character(rep(c(1:10), each=3)), readout=log10(as.numeric(unlist(x[3:32]))), severe=rep(c(0,1), each=15))
}

genes1=c("CXCL9", "CXCL10",  "CXCL11"); names(genes1)=genes1
rows=c("CXCL9"=list(4:8), "CXCL10"=list(12:16), "CXCL11"=list(20:24))

lapply(genes1, function (gene) {
  res=lapply(rows[[gene]], function (i) doOneRow(dat1[i-1,]))
  names(res)=doses
  res
}) -> dat.list

mysapply(genes1, function (gene) {
  fits <- lapply(doses, function(d) {
  
    fit <- lmer(readout ~ severe + (1|donor), data=dat.list[[gene]][[d]])
  
    # flag warning if singular or has convergence messages
    warn_flag <- FALSE
  
    # singularity
    if (isSingular(fit, tol = 1e-4)) {
      warn_flag <- TRUE
    }
  
    # convergence warnings stored in optinfo
    msgs <- fit@optinfo$conv$lme4$messages
    if (!is.null(msgs)) {
      warn_flag <- TRUE
    }
  
    # store attribute
    attr(fit, "warning_occurred") <- warn_flag
    attr(fit, "convergence_messages") <- msgs
  
    fit
  })
  
  
  names(fits)=doses
  tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
  res = tab[2,]
  
  # check for warnings
  warnings = sapply(fits, function(fit) attr(fit, "warning_occurred"))
  res[warnings] = paste0(res[warnings], "W")

  sapply(doses, function(d) len(unique(dat.list[[gene]][[d]]$readout))==1) -> all_same
  res[all_same] = "all values identical"
  res
}) -> tab

tab=cbind(rownames(tab), tab)
colnames(tab)[1]="7B"
kable(tab, align="c", row.names=F, caption="Linear mixed models.")%>%  
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(1, width = "1.1cm") %>%
  column_spec(2:(ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")


```


```{r Fig 7B rep, echo=F, message=T, results='asis'}

mysapply(genes1, function (gene) {
  # fits = lapply(doses, function(d) {
  #   # suppressWarnings(
  #     lmer(readout ~ severe + (1|donor), data=dat.list[[gene]][[d]])
  #   # )
  # })
  # names(fits)=doses
  # tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
  # res = tab[2,]
  
  # unpaired t test
  out = sapply(doses, function(d) {
    tmp1=dat.list[[gene]][[d]] %>%
      group_by(donor, severe) %>%
      summarize(readout=mean(readout), .groups="drop")
    t.test(readout~severe, data=tmp1)$p.val
  })
  names(out)=doses
  res = formatPvalues(out)
  
  sapply(doses, function(d) len(unique(dat.list[[gene]][[d]]$readout))==1) -> all_same
  res[all_same] = "all values identical"
  res
}) -> tab

tab=cbind(rownames(tab), tab)
colnames(tab)[1]="7B"
kable(tab, align="c", row.names=F, caption="Unpaired t tests.")%>%  
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(1, width = "1.1cm") %>%
  column_spec(2:(ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")


```



\newpage
## Fig 8

Data is log10 transformed.

**Question**: Determine if there is a difference in pg/mL of protein between asymptomatic and severe in each treatment group (mock, 100U, virus, 100U+virus). 


```{r Fig 8 q1, echo=F, message=T, results='asis'}
dat1=suppressMessages(
  readxl::read_xlsx("data/Data_CytokinePaper_IRH-AZ_2025-11-6.xlsx", sheet=glue("Figure 8"))
)

doOneRow=function (x) {
  data.frame(donor=as.character(rep(c(1:10), each=3)), readout=log10(as.numeric(unlist(x[-(1)]))), severe=rep(c(0,1), each=15))
}

genes=c("CXCL9 4HPT", "CXCL9 12HPT",  "CXCL10 4HPT", "CXCL10 12HPT"); names(genes)=genes
rows=c("CXCL9 4HPT"=list(5:8), "CXCL9 12HPT"=list(12:15), "CXCL10 4HPT"=list(19:22), "CXCL10 12HPT"=list(26:29))

lapply(genes, function (gene) {
  res=lapply(rows[[gene]], function (i) doOneRow(dat1[i-1,]))
  names(res)=trts
  res
}) -> dat.list


mysapply(genes, function (gene) {
  res = sapply(trts, function(d) {
    if (len(unique(dat.list[[gene]][[d]]$readout))==1) {
      return(NA)
    } else {
      fits=list(lmer(readout ~ severe + (1|donor), data=dat.list[[gene]][[d]]))
      tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
      out=tab[2,]
      
      warn_flag=F
      
      # singularity
      if (isSingular(fits[[1]], tol = 1e-4)) {
        warn_flag <- TRUE
      }
    
      # convergence warnings stored in optinfo
      msgs <- fits[[1]]@optinfo$conv$lme4$messages
      if (!is.null(msgs)) {
        warn_flag <- TRUE
      }
      
      if (warn_flag) {
        out = paste0(out, "W")
      }
      
      out
    }
  })
  names(res)=trts
  sapply(trts, function(d) len(unique(dat.list[[gene]][[d]]$readout))==1) -> all_same
  res[all_same] = "all values identical"
  res
}) -> tab

kable(tab, align="c", caption="Linear mixed models.")%>%  
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(1, width = "1.1cm") %>%  
  column_spec(2:(1+ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")


# now do it with unpaired t test
mysapply(genes, function (gene) {
  res = sapply(trts, function(d) {
    if (len(unique(dat.list[[gene]][[d]]$readout))==1) {
      return(NA)
    } else {
      dat.tmp = dat.list[[gene]][[d]] %>%
        group_by(donor, severe) %>%
        summarize(readout=mean(readout), .groups="drop")
      formatPvalues(t.test(readout~severe, data=dat.tmp)$p.val) 
    }
  })
  names(res)=trts
  sapply(trts, function(d) len(unique(dat.list[[gene]][[d]]$readout))==1) -> all_same
  res[all_same] = "all values identical"
  res
}) -> tab

kable(tab, align="c", caption="Unpaired t test pvals.")%>%  
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(1, width = "1.1cm") %>%  
  column_spec(2:(1+ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")

```

\newpage
**Question**: Determine if there is a difference in pg/mL of protein between Mock and each treatment (100U, Virus, 100U+Virus) within each group (asymptomatic, severe). For example, if there is a statistical difference between asymptomatic mock vs asymptomatic 100U

```{r Fig 8 q2, echo=F, message=T, results='asis'}
for (severe in 0:1) {
    
  mysapply(genes, function (gene) {
    res = sapply(trts[-1], function(d) {
      dat.tmp = rbind(
        cbind(trt="Mock", dat.list[[gene]][["Mock"]]),
        cbind(trt=d, dat.list[[gene]][[d]])
      )
      dat.tmp$trt=factor(dat.tmp$trt, levels=trts)
      
      if (severe==0) {
        dat.tmp = subset(dat.tmp, severe==0)
      } else {
        dat.tmp = subset(dat.tmp, severe==1)
      }
      
      if (len(unique(dat.tmp$readout))==1) {
        return("all values identical")
      } else {
        # # lmm
        # getFormattedSummary(list(lmer(readout ~ trt + (1|donor), data=dat.tmp)), robust=F, p.digits=3, trunc.large.est = F)[2,]
        
        # paired t test
        formatPvalues(f.paired.t(dat.tmp, "readout", "trt"))
      }
    })
    res
  }) -> tab
  colnames(tab)=glue("{colnames(tab)} vs. Mock") 
  
  print(kable(tab, align="c", caption=glue("Paired t test p values for {ifelse(severe==0, 'asym', 'severe')}."))%>%  
    kable_styling(latex_options = "hold_position") %>% 
    column_spec(2, width = "1.1cm") %>%  
    column_spec(1:(1+ncol(tab)), width = "3cm")%>%  
    row_spec(0, bold = TRUE, align = "c")%>%
    row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]"))
  
}

```

\newpage

# Cellprofiler data analysis

Data is not log10 transformed.

Median of cellwise integrated intensity per donor per replicate is used as MFI.


```{r cellprofiler data, echo=F, message=T, results='asis'}
dat1=readxl::read_xlsx("data/2025-11-6_pSTAT1_NuclearFlr_Data.xlsx",
                        col_types = c("text", "numeric", "text", "numeric", "numeric", "numeric"),
                        sheet="Replicate_1")
dat2=readxl::read_xlsx("data/2025-11-6_pSTAT1_NuclearFlr_Data.xlsx",
                        col_types = c("text", "numeric", "text", "numeric", "numeric", "numeric"),
                        sheet="Replicate_2")

names(dat2) <- names(dat1)
rbind(cbind(replicate=1, dat1), cbind(replicate=2, dat2)) -> dat.cellprofiler
names(dat.cellprofiler)[names(dat.cellprofiler)=="Integrated Intensity"]="Integrated_Intensity"
names(dat.cellprofiler)[names(dat.cellprofiler)=="Donor"]="donor"

# compute MFI, defined as median of Integrated_Intensity, for each Sample, Replicate, Donor combination
dat.cellprofiler %>%
  group_by(Sample, replicate, donor) %>%
  summarize(MFI=median(Integrated_Intensity, na.rm=TRUE), .groups="drop") -> dat.mfi

# filter out donor NHEK
dat.mfi = subset(dat.mfi, donor != "NHEK")

# add severe column
dat.mfi$severe = ifelse(startsWith(dat.mfi$Sample, "Severe"), 1, 0)
# add dose column
dat.mfi$Sample = sub("^(Asym|Severe)_", "", dat.mfi$Sample)

strsplit(dat.mfi$Sample, "_") -> sample_parts
dat.mfi$dose = sapply(sample_parts, function(x) x[1])
dat.mfi$time = sapply(sample_parts, function(x) x[2])

# change dose "M" to "0", "01" to "0.1"
dat.mfi$dose[dat.mfi$dose=="M"]="0"
dat.mfi$dose[dat.mfi$dose=="01"]="0.1"
dat.mfi$dose = factor(dat.mfi$dose, levels=doses)

dat.mfi$time = factor(dat.mfi$time, levels=times2)

# if not converted to data frame, an error is thrown while knitting but not when run interactively
mywrite.csv(as.data.frame(dat.mfi), "data/Cellprofiler_MFI_data")

# myboxplot(MFI~severe, subset(dat.mfi, time=="120M" & dose=="10"), col.points=as.numeric(dat.mfi$donor), xlab="Severe", cex=1.2, pch=16)

```


**Question 1**: Determine if there are statistically significant differences between each time (30, 60, 120, 240 minutes) at each dose (0, 0.1, 1, 10, 100). Data represents fluorescence intensity in individual cells, collected across two independent experiments. Data is pooled across 10. For example, comparing 100U at 30 minutes to 100U at 60 minutes. 


```{r cellprofiler q1, echo=F, message=T, results='asis'}

lapply(doses, function (.dose) {
  res=lapply(times2, function (.t) {
    subset(dat.mfi, dose==.dose & time==.t)
  })
  res
}) -> dat.list

mysapply(doses, function (d) {
  # # lmm
  # fits = lapply(2:len(times2), function(i) {
  #   dat.tmp = rbind(
  #     cbind(time=times2[i-1], dat.list[[d]][[times2[i-1]]], row.names=NULL),
  #     cbind(time=times2[i]  , dat.list[[d]][[times2[i]]], row.names=NULL)
  #   )
  #   # suppressWarnings(
  #     lmer(MFI ~ time + (1|donor), data=dat.tmp)
  #   # )
  # })
  # names(fits)=times2[-1]
  # tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
  # res = tab[2,]
  # res
  
  # paired t test
  out = sapply(2:len(times2), function(i) {
    dat.tmp = rbind(
      cbind(dat.list[[d]][[times2[i-1]]], row.names=NULL),
      cbind(dat.list[[d]][[times2[i]]], row.names=NULL)
    )
    dat.tmp$time=factor(dat.tmp$time, levels=times2)
    f.paired.t(dat.tmp, "MFI", "time")
  })
  names(out)=paste0(times2[2:length(times2)], " vs. ", times2[2:length(times2)-1])
  res = formatPvalues(out)
  res
}) -> tab
colnames(tab)=glue("{colnames(tab)} vs. Prev. Time") 

kable(tab, align="c", caption="Paired t test p values.")%>%  
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(1, width = "1.1cm") %>%  
  column_spec(2:(1+ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")


```

\newpage

**Question 2**: Determine if there is a statistically significant difference between each dose (0.1, 1, 10, 100) to 0 at 30 minutes. For example, comparing 0 to 0.1, 0 to 1, 0 to 10, and 0 to 100 at 30 minutes.

```{r cellprofiler q2, echo=F, message=T, results='asis'}

lapply(doses, function (.dose) {
  res=lapply(times2, function (.t) {
    subset(dat.mfi, dose==.dose & time==.t)
  })
  res
}) -> dat.list


mysapply(times2, function (t) {
  # # lmm
  # fits = lapply(doses[-1], function(d) {
  #   dat.tmp = rbind(
  #     cbind(dose="0", dat.list[["0"]][[t]]),
  #     cbind(dose=d,   dat.list[[d]][[t]])
  #   )
  #   # suppressWarnings(
  #     lmer(MFI ~ dose + (1|donor), data=dat.tmp)
  #   # )
  # })
  # names(fits)=doses[-1]
  # tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
  # res = tab[2,]
  # res
  
  # paired t test
  out = sapply(doses[-1], function(d) {
    dat.tmp = rbind(
      cbind(dat.list[["0"]][[t]], row.names=NULL),
      cbind(dat.list[[d]][[t]], row.names=NULL)
    )
    dat.tmp$dose=factor(dat.tmp$dose, levels=doses)
    f.paired.t(dat.tmp, "MFI", "dose")
  })
  names(out)=paste0(doses[2:length(doses)], " vs. 0")
  res = formatPvalues(out)
  res
  
}) -> tab
# colnames(tab)=glue("{colnames(tab)} vs. 0") 

kable(tab, align="c", caption="Paired t test")%>%  
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(1, width = "1.1cm") %>%  
  column_spec(2:(1+ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")


```

\newpage

**Question 3**: Determine if there are any statistically significant differences between asymptomatic and severe at each dose (0, 0.1, 1, 10, 100) and each time (30, 60, 120, 240 minutes). For example, if there is a significant difference between asymptomatic 100U and severe 100U at 30 minutes

```{r cell profiler q3, echo=F, message=T, results='asis'}

lapply(doses, function (.dose) {
  res=lapply(times2, function (.t) {
    subset(dat.mfi, dose==.dose & time==.t)
  })
  res
}) -> dat.list

mysapply(doses, function (dose) {
  
  fits <- lapply(times2, function(t) {
  
    fit <- lmer(MFI ~ severe + (1|donor), data=dat.list[[dose]][[t]])
  
    # flag warning if singular or has convergence messages
    warn_flag <- FALSE
  
    # singularity
    if (isSingular(fit, tol = 1e-4)) {
      warn_flag <- TRUE
    }
  
    # convergence warnings stored in optinfo
    msgs <- fit@optinfo$conv$lme4$messages
    if (!is.null(msgs)) {
      warn_flag <- TRUE
    }
  
    # store attribute
    attr(fit, "warning_occurred") <- warn_flag
    attr(fit, "convergence_messages") <- msgs
  
    fit
  })
  
  names(fits)=times2
  tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
  res = tab[2,]
  
  # check for warnings
  warnings = sapply(fits, function(fit) attr(fit, "warning_occurred"))
  res[warnings] = paste0(res[warnings], "W")
  
  res
}) -> tab

kable(tab, align="c", caption="Linear mixed models.")%>%  
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(1, width = "1.1cm") %>%  
  column_spec(2:(1+ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")


```

```{r cell profiler q3 t test, echo=F, message=T, results='asis'}

mysapply(doses, function (dose) {
  pvals = sapply(times2, function(t) {
    tmp1=dat.list[[dose]][[t]] %>%
      group_by(donor, severe) %>%
      summarize(MFI=mean(MFI), .groups="drop") 
    t.test(MFI~severe, data=tmp1)$p.val
  })
  formatPvalues(pvals)
}) -> tab

kable(tab, align="c", caption="Unpaired t tests.")%>%  
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(1, width = "1.1cm") %>%  
  column_spec(2:(1+ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[2pt]")

```

\newpage
# Appendix
```{r, echo = FALSE, message = FALSE, warning = FALSE, results='asis'}
commit_hash <- system("git rev-parse HEAD", intern = TRUE)
git_url <- sub("\\.git$", paste0("/commits/", commit_hash), system("git remote get-url origin", intern = TRUE))
cat("This report was built with code from [", sprintf("[this commit](%s)", git_url), "].", sep="")
```    