---
title: "Analysis Leaving Out Donor 8"
date: "`r Sys.Date()`"
output: pdf_document
header-includes:
  - \usepackage{array}

---

```{r setup, include=FALSE}
library(kyotil)
library(knitr)
library(kableExtra)
library(lme4)
library(readxl)
library(glue)
library(dplyr)

# global variables, don't redefine in each section
cytokines=c("CXCL10", "CCL5",  "IL15",  "IL12A", "IFI16"); names(cytokines)=cytokines
trts=c("Mock", "100U", "Virus", "Virus+100U"); names(trts)=trts
donors=c(4,5,7,10,1,2,3,6,8,9)
doses=c("0", "0.1", "1", "10", "100"); names(doses)=doses
tissues=c("Keratinocytes", "Fibroblasts"); names(tissues)=tissues
times=c("Untreated", "4HPT", "24HPT"); names(times)=times
times2=c("30M", "60M", "120M", "240M"); names(times2)=times2


f.paired.t = function(dat, outcome, cond, print=F) {

  # take average of replicates for each donor and trt
  tmp = dat %>%
    group_by(donor, !!sym(cond)) %>%
    summarize(fc = mean(.data[[outcome]], na.rm = TRUE), .groups = "drop")

  # reshape to wide format with one column per trt
  tmp1 = reshape(as.data.frame(tmp), idvar = "donor", timevar = cond,
                 direction = "wide")

  # extract the two treatment groups inside the cond column
  groups <- unique(dat[[cond]])
  if (length(groups) != 2) {
    stop("The cond column must contain exactly 2 treatment levels.")
  }
  # rename for clarity
  g1 <- groups[1]
  g2 <- groups[2]

  # print(tmp1)
  
  # run paired t-test
  test = t.test(tmp1[["fc."%.%g1]], tmp1[["fc."%.%g2]], paired = TRUE)

  if (print) {
    myboxplot(list(tmp1[["fc."%.%g1]], tmp1[["fc."%.%g2]]))
    print(tmp1)
    print(test)
  }
  
  test$p.value
}

```

\clearpage

## Fig 6A, 6B, and 6C

Data is log10 transformed for 6B and 6C and not 6A.

**Question**: Determine if there is a statistically significant difference between asymptomatic and severe donors at each dose (6A, 6B) or at each time (6C) for each gene (CXCL10, CCL5, IL15, IL12, IFI16) at dose 100. 

Data represents three independent biological replicates for all 10 donors. 

We first analyzed the data with linear mixed models. There are some warnings about singular values in 6B and 6C. Hence we re-analyzed the data with unpaired t tests.



```{r Fig 6ABC, echo=F, message=T, results='asis'}

genes=c("CXCL10", "CCL5",  "IL15",  "IL12A", "IFI16"); names(genes)=genes

for (a in c("A", "B", "C")) {
  
  dat1=suppressMessages(
    readxl::read_xlsx("data/Data_CytokinePaper_IRH-AZ_2025-11-6.xlsx", sheet=glue("Figure 6{a}"))
  )
  
  doOneRow=function (x) {
    res = data.frame(donor=as.character(rep(c(1:10), each=3)), 
               readout=(as.numeric(unlist(x[-(1:2)]))), 
               severe=rep(c(0,1), each=15))
    if (a=="B" | a=="C") {
      res$readout = log10(res$readout)
    }
    # filter out donor 8
    res = subset(res, donor!="8")
    res
  }
  
  if (a=="A" | a=="B") {
    rows=c(CXCL10=list(4:8), CCL5=list(12:16), IL15=list(20:24), IL12A=list(28:32), IFI16=list(36:40))
  } else if (a=="C") {
    rows=c(CXCL10=list(4:6), CCL5=list(10:12), IL15=list(16:18), IL12A=list(22:24), IFI16=list(28:30))
  }
  
  tmp = if (a=="A" | a=="B") doses else times
  
  lapply(genes, function (gene) {
    res=lapply(rows[[gene]], function (i) doOneRow(dat1[i-1,]))
    names(res)= tmp
    res
  }) -> dat.list
  
  for (iter in 1:2) {
    mysapply(genes, function (gene) {
      # print(gene)
      if (iter==1) {
        # lmm
        fits <- lapply(tmp, function(d) {
        
          fit <- lmer(readout ~ severe + (1|donor), data = dat.list[[gene]][[d]])
        
          # flag warning if singular or has convergence messages
          warn_flag <- FALSE
        
          # singularity
          if (isSingular(fit, tol = 1e-4)) {
            warn_flag <- TRUE
          }
        
          # convergence warnings stored in optinfo
          msgs <- fit@optinfo$conv$lme4$messages
          if (!is.null(msgs)) {
            warn_flag <- TRUE
          }
        
          # store attribute
          attr(fit, "warning_occurred") <- warn_flag
          attr(fit, "convergence_messages") <- msgs
        
          fit
        })

        tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
        out = tab[2,]
        # check for warnings
        warnings = sapply(fits, function(fit) attr(fit, "warning_occurred"))
        # print(warnings)
        out[warnings] = paste0(out[warnings], "W")
        out
        
      } else {
        # unpaired t test
        pvals = sapply(tmp, function(d) {
          tmp1=dat.list[[gene]][[d]] %>%
            group_by(donor, severe) %>%
            summarize(readout=mean(readout), .groups="drop")
          # par(mfrow=c(1,2))
          # myboxplot(readout~severe, data=tmp1, ylim=c(-1,1))
          # myboxplot(readout~severe, data=tmp1)
          t.test(readout~severe, data=tmp1)$p.val
        })
        formatPvalues(pvals)
      }  
    }) -> tab
    tab
  
    tab=cbind(rownames(tab), tab)
    colnames(tab)[1]=glue("6{a}")
    print(
      kable(tab, align="c", row.names=FALSE, caption=ifelse(iter==1,"Linear mixed models.", "Unpaired t tests.")) %>%  
      kable_styling(latex_options = "hold_position") %>% 
      column_spec(1, width = "1.1cm") %>%  
      column_spec(2:(ncol(tab)), width = "3cm")%>%  # no 1+ncol b/c row.names is FALSE
      row_spec(0, bold = TRUE, align = "c")%>%
      row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")
    ) 
  }
  
  cat("\\newpage\n")   # page break in LaTeX

}
```

## Fig 7A and 7B


**Question**: Determine if there is a statistically significant difference between the expression level of each gene (CXCL10, CCL5, IL15, IL12, IFI16) at each dose compared to the previous dose. For clarity, this is comparing 0.0 to 0.1, 0.1 to 1, 1 to 10, and 10 to 100. Data represents three independent biological replicates from 10 different donors

Data is log10 transformed.

```{r Fig 7A, echo=F, message=T, results='asis'}

dat1=suppressMessages(
  readxl::read_xlsx("data/Data_CytokinePaper_IRH-AZ_2025-11-6.xlsx", sheet=glue("Figure 7A"))
)

doOneRow=function (x) {
  # use x[3:32] instead of x[-(1:2)] b/c there are some extra columns, not visible in excel, at the end
  res = data.frame(donor=as.character(rep(c(1:10), each=3)), readout=log10(as.numeric(unlist(x[3:32]))), severe=rep(c(0,1), each=15))
  # filter out donor 8
  res = subset(res, donor!="8")
  res
}

genes1=c("CXCL9", "CXCL10",  "CXCL11"); names(genes1)=genes1
rows=c("CXCL9"=list(4:8), "CXCL10"=list(12:16), "CXCL11"=list(20:24))

lapply(genes1, function (gene) {
  res=lapply(rows[[gene]], function (i) doOneRow(dat1[i-1,]))
  names(res)=doses
  res
}) -> dat.list

mysapply(genes1, function (gene) {
  # # linear mixed model
  # fits = lapply(2:len(doses), function(i) {
  #   dat.tmp = rbind(
  #     cbind(dose=doses[i-1], dat.list[[gene]][[doses[i-1]]], row.names=NULL),
  #     cbind(dose=doses[i],   dat.list[[gene]][[doses[i]]], row.names=NULL)
  #   )
  #   dat.tmp$dose=factor(dat.tmp$dose, levels=doses)
  # 
  #   # suppressWarnings(
  #     lmer(readout ~ dose + (1|donor), data=dat.tmp)
  #   # )
  # })
  # names(fits)=doses[-1]
  # tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
  # res = tab[2,]
  
  # paired t test
  out = sapply(2:len(doses), function(i) {
    dat.tmp = rbind(
      cbind(dose=doses[i-1], dat.list[[gene]][[doses[i-1]]], row.names=NULL),
      cbind(dose=doses[i],   dat.list[[gene]][[doses[i]]], row.names=NULL)
    )
    dat.tmp$dose=factor(dat.tmp$dose, levels=doses)
    f.paired.t(dat.tmp, "readout", "dose")
  })
  names(out)=paste0(doses[2:length(doses)], " vs. ", doses[2:length(doses)-1])
  res = formatPvalues(out)
    
  all_same = sapply(2:len(doses), function(i) {
    dat.tmp = rbind(
      cbind(dose=doses[i-1], dat.list[[gene]][[doses[i-1]]], row.names=NULL),
      cbind(dose=doses[i],   dat.list[[gene]][[doses[i]]], row.names=NULL)
    )
    len(unique(dat.tmp$readout))==1
  })
  res[all_same] = "all values identical"
  res
}) -> tab
# colnames(tab)=glue("{colnames(tab)} vs. Prev. Dose")

tab=cbind(rownames(tab), tab)
colnames(tab)[1]="7A"
kable(tab, align="c", caption="Paired t tests.", row.names=F)%>%  
  kable_styling(latex_options = "hold_position") %>% 
  # column_spec(1, width = "1.1cm") %>%  
  column_spec(1:(ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")


```

\newpage

**Question**: Determine if there is a statistically significant difference between asymptomatic and severe donors at each dose (0, 0.1, 1, 10, 100). Data represents three independent biological replicates for all 10 donors.

Data is log10 transformed.

```{r Fig 7B, echo=F, message=T, results='asis'}

dat1=suppressMessages(
  readxl::read_xlsx("data/Data_CytokinePaper_IRH-AZ_2025-11-6.xlsx", sheet=glue("Figure 7B"))
)

doOneRow=function (x) {
  # use x[3:32] instead of x[-(1:2)] b/c there are some extra columns, not visible in excel, at the end
  res = data.frame(donor=as.character(rep(c(1:10), each=3)), readout=log10(as.numeric(unlist(x[3:32]))), severe=rep(c(0,1), each=15))
  # filter out donor 8
  res = subset(res, donor!="8")
  res
}

genes1=c("CXCL9", "CXCL10",  "CXCL11"); names(genes1)=genes1
rows=c("CXCL9"=list(4:8), "CXCL10"=list(12:16), "CXCL11"=list(20:24))

lapply(genes1, function (gene) {
  res=lapply(rows[[gene]], function (i) doOneRow(dat1[i-1,]))
  names(res)=doses
  res
}) -> dat.list

mysapply(genes1, function (gene) {
  fits <- lapply(doses, function(d) {
  
    fit <- lmer(readout ~ severe + (1|donor), data=dat.list[[gene]][[d]])
  
    # flag warning if singular or has convergence messages
    warn_flag <- FALSE
  
    # singularity
    if (isSingular(fit, tol = 1e-4)) {
      warn_flag <- TRUE
    }
  
    # convergence warnings stored in optinfo
    msgs <- fit@optinfo$conv$lme4$messages
    if (!is.null(msgs)) {
      warn_flag <- TRUE
    }
  
    # store attribute
    attr(fit, "warning_occurred") <- warn_flag
    attr(fit, "convergence_messages") <- msgs
  
    fit
  })
  
  
  names(fits)=doses
  tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
  res = tab[2,]
  
  # check for warnings
  warnings = sapply(fits, function(fit) attr(fit, "warning_occurred"))
  res[warnings] = paste0(res[warnings], "W")

  sapply(doses, function(d) len(unique(dat.list[[gene]][[d]]$readout))==1) -> all_same
  res[all_same] = "all values identical"
  res
}) -> tab

tab=cbind(rownames(tab), tab)
colnames(tab)[1]="7B"
kable(tab, align="c", row.names=F, caption="Linear mixed models.")%>%  
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(1, width = "1.1cm") %>%
  column_spec(2:(ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")


```


```{r Fig 7B rep, echo=F, message=T, results='asis'}

mysapply(genes1, function (gene) {
  # fits = lapply(doses, function(d) {
  #   # suppressWarnings(
  #     lmer(readout ~ severe + (1|donor), data=dat.list[[gene]][[d]])
  #   # )
  # })
  # names(fits)=doses
  # tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
  # res = tab[2,]
  
  # unpaired t test
  out = sapply(doses, function(d) {
    tmp1=dat.list[[gene]][[d]] %>%
      group_by(donor, severe) %>%
      summarize(readout=mean(readout), .groups="drop")
    t.test(readout~severe, data=tmp1)$p.val
  })
  names(out)=doses
  res = formatPvalues(out)
  
  sapply(doses, function(d) len(unique(dat.list[[gene]][[d]]$readout))==1) -> all_same
  res[all_same] = "all values identical"
  res
}) -> tab

tab=cbind(rownames(tab), tab)
colnames(tab)[1]="7B"
kable(tab, align="c", row.names=F, caption="Unpaired t tests.")%>%  
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(1, width = "1.1cm") %>%
  column_spec(2:(ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")


```



\newpage
# Updated Luminex data

Data is log10 transformed.

**Question**: Is there a difference between asymptomatic and severe donors at each dose in each cytokine? IE 0U vs 0U, 1U vs 1U, 100U vs 100U.

```{r Fig updated Luminex q1, echo=F, message=T, results='asis'}
dat1=suppressMessages(
  readxl::read_xlsx("data/UpdatedLuminex_Data.xlsx", sheet=glue("Sheet1"))
)

doOneRow=function (x) {
  res = data.frame(donor=as.character(rep(c(1:10), each=3)), readout=log10(as.numeric(unlist(x[-(1)]))), severe=rep(c(0,1), each=15))
  # filter out donor 8
  res = subset(res, donor!="8")
  res
}

genes=c("CXCL9", "CXCL10", "CXCL11"); names(genes)=genes
rows=c("CXCL9"=list(5:7), "CXCL10"=list(15:17), "CXCL11"=list(25:27))

rownames=c("0U", "1U", "100U")

lapply(genes, function (gene) {
  res=lapply(rows[[gene]], function (i) doOneRow(dat1[i-1,]))
  names(res)=rownames
  res
}) -> dat.list


mysapply(genes, function (gene) {
  res = sapply(rownames, function(d) {
    if (len(unique(dat.list[[gene]][[d]]$readout))==1) {
      return(NA)
    } else {
      fits=list(lmer(readout ~ severe + (1|donor), data=dat.list[[gene]][[d]]))
      tab=getFormattedSummary(fits, robust=F, p.digits=3, trunc.large.est = F)
      out=tab[2,]
      
      warn_flag=F
      
      # singularity
      if (isSingular(fits[[1]], tol = 1e-4)) {
        warn_flag <- TRUE
      }
    
      # convergence warnings stored in optinfo
      msgs <- fits[[1]]@optinfo$conv$lme4$messages
      if (!is.null(msgs)) {
        warn_flag <- TRUE
      }
      
      if (warn_flag) {
        out = paste0(out, "W")
      }
      
      out
    }
  })
  names(res)=rownames
  sapply(rownames, function(d) len(unique(dat.list[[gene]][[d]]$readout))==1) -> all_same
  res[all_same] = "all values identical"
  res
}) -> tab

# set test p value to N/A for CXCL9 and CXCL11 at 0U because all values are under LOD and they only differ in concentration factors
tab["CXCL9","0U"] <- tab["CXCL11","0U"] <- NA

kable(tab, align="c", caption="Linear mixed models. P values set to N/A for CXCL9 and CXCL11 at 0U because all values are under LOD and they only differ inconcentration factors.")%>%  
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(1, width = "1.1cm") %>%  
  column_spec(2:(1+ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")


# now do it with unpaired t test
mysapply(genes, function (gene) {
  res = sapply(rownames, function(d) {
    if (len(unique(dat.list[[gene]][[d]]$readout))==1) {
      return(NA)
    } else {
      dat.tmp = dat.list[[gene]][[d]] %>%
        group_by(donor, severe) %>%
        summarize(readout=mean(readout), .groups="drop")
      formatPvalues(t.test(readout~severe, data=dat.tmp)$p.val) 
    }
  })
  names(res)=rownames
  sapply(rownames, function(d) len(unique(dat.list[[gene]][[d]]$readout))==1) -> all_same
  res[all_same] = "all values identical"
  res
}) -> tab

# set test p value to N/A for CXCL9 and CXCL11 at 0U because all values are under LOD and they only differ in concentration factors
tab["CXCL9","0U"] <- tab["CXCL11","0U"] <- NA

kable(tab, align="c", caption="Unpaired t test pvals. P values set to N/A for CXCL9 and CXCL11 at 0U because all values are under LOD and they only differ inconcentration factors.")%>%  
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(1, width = "1.1cm") %>%  
  column_spec(2:(1+ncol(tab)), width = "3cm")%>%  
  row_spec(0, bold = TRUE, align = "c")%>%
  row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]")

```

\newpage
**Question**: Is there a difference between 0U and 1U for each chemokine? Between 0U and 100U? Between 1U and 100U?

```{r updated Luminex q2, echo=F, message=T, results='asis'}
for (severe in 0:1) {
    
  mysapply(genes, function (gene) {
    res = sapply(1:3, function(iter) {
      if (iter==1) {
        d1 = "0U"; d2 = "1U"
      } else if (iter==2) {
        d1 = "0U"; d2 = "100U"
      } else if (iter==3) {
        d1 = "1U"; d2 = "100U"
      }
      dat.tmp = rbind(
        cbind(trt=d1, dat.list[[gene]][[d1]]),
        cbind(trt=d2, dat.list[[gene]][[d2]])
      )
      dat.tmp$trt=factor(dat.tmp$trt, levels=rownames)
      
      if (severe==0) {
        dat.tmp = subset(dat.tmp, severe==0)
      } else {
        dat.tmp = subset(dat.tmp, severe==1)
      }
      
      if (len(unique(dat.tmp$readout))==1) {
        return("all values identical")
      } else {
        # # lmm
        # getFormattedSummary(list(lmer(readout ~ trt + (1|donor), data=dat.tmp)), robust=F, p.digits=3, trunc.large.est = F)[2,]
        
        # paired t test
        v = formatPvalues(f.paired.t(dat.tmp, "readout", "trt"))
        names(v) = glue("{d1} vs. {d2}")
        v
      }
    })
    res
  }) -> tab
  
  print(kable(tab, align="c", caption=glue("Paired t test p values for {ifelse(severe==0, 'asym', 'severe')}."))%>%  
    kable_styling(latex_options = "hold_position") %>% 
    column_spec(2, width = "1.1cm") %>%  
    column_spec(1:(1+ncol(tab)), width = "3cm")%>%  
    row_spec(0, bold = TRUE, align = "c")%>%
    row_spec(1:nrow(tab), extra_latex_after = "\\addlinespace[6pt]"))
  
}

```



\newpage
